\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2024-09-21 Manuscript/manuscript.tex
adjusted flow in intro

The first objective of this paper is to investigate whether reaction
times at the 2022 World Championships were significantly different
from other competitions, focusing specifically on athletes who
competed in multiple events. By employing a matched-pairs design, we
compare reaction times from the same athletes across different
championships, including the 2022, 2019, and 2023 World Championships,
as well as the 2022 national-level competitions. This approach allows
us to isolate the effect of the competition year on reaction times
while controlling for individual athlete performance. Our analysis
provide a more formal statistical evaluation of the disqualification 
threshold, building upon the informal comparisons previously made 
by others.

There is an ongoing debate over the appropriateness of the 0.1-second
threshold for disqualification, with past studies suggesting both
lower and higher limits. \citet{pain2007sprint} examined 
reaction times in nine male athletes and found that the fastest athlete 
had an average reaction time of 0.087 seconds, with a standard deviation 
of 0.004 seconds. Similarly, \citet{komi2009iaaf}, in a study commissioned 
by the IAAF, recommended lowering the disqualification threshold to 0.08 
or 0.085 seconds, arguing that athletes are capable of faster reaction 
times than the current threshold allows. On the other hand, some 
researchers have argued for increasing the threshold.
\citet{brosnan2017effects} suggested raising it to 0.115 seconds based
on data from European and World Championships, while a study of the
2008 Olympics supported the conclusion that athletes cannot reliably
react within 0.1 seconds \citep{lipps2011implications}. Interestingly,
\citet{brosnan2017effects} and \citet{lipps2011implications} relied on
retrospective event data, whereas \citet{pain2007sprint} and
\citet{komi2009iaaf} conducted controlled experiments to assess
reaction time capabilities. The IAAF has not revised the threshold
since its implementation in 1999.

The second objective of this paper is to further investigate the
appropriateness of the 0.1-second threshold by modeling reaction times
from athletes who competed in world championships from 1999 onward. We
employ a gamma linear mixed-effects model with two random effects: one
for the venue (year) and another for the heat nested within each
venue. This model allows us to estimate the probability of an
athlete’s reaction time falling below a specific threshold, such as
0.1 seconds. By examining reaction times across different years and
heats, we can assess the consistency of reaction times and whether the
current disqualification threshold is supported by statistical
evidence.

>>>>

The 0.1-second disqualification threshold in track events has been 
a subject of ongoing debate, with researchers presenting evidence 
both in favor of lowering and raising the
limit. \citet{pain2007sprint} examined
reaction times in nine male athletes and found that the fastest athlete 
had an average reaction time of 0.087 seconds, with a standard deviation 
of 0.004 seconds. Similarly, \citet{komi2009iaaf}, in a study commissioned 
by the IAAF, recommended lowering the disqualification threshold to 0.08 
or 0.085 seconds, arguing that athletes are capable of faster reaction 
times than the current threshold allows. On the other hand, some 
researchers have argued for increasing the threshold.
\citet{brosnan2017effects} suggested raising it to 0.115 seconds based
on data from European and World Championships, while a study of the
2008 Olympics supported the conclusion that athletes cannot reliably
react within 0.1 seconds \citep{lipps2011implications}. Interestingly,
\citet{brosnan2017effects} and \citet{lipps2011implications} relied on
retrospective event data, whereas \citet{pain2007sprint} and
\citet{komi2009iaaf} conducted controlled experiments to assess
reaction time capabilities. The IAAF has not revised the threshold
since its implementation in 1999.

This paper has two primary objectives. First, we investigate whether 
reaction times at the 2022 World Championships were significantly 
different from other competitions, focusing on athletes who competed 
in multiple events. Using a matched-pairs design, we compare reaction 
times across the 2022, 2019, and 2023 World Championships, as well as 
the 2022 national-level competitions. This approach isolates the 
effect of the competition year while controlling for individual 
performance. The second objective is to assess the 
appropriateness of the 0.1-second threshold by modeling reaction 
times from World Championships held from 1999 onward. We use a gamma 
linear mixed-effects model with random effects for both venue and heat. 
This model estimates the probability of reaction times falling below 
the threshold, helping evaluate the consistency of reaction times and 
whether the current threshold is statistically justified.
